Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-03-11 main.Rmd
content cut till cyclic calendar algebra

The first condition in Definition \ref{def:linear} implies that the granules in a linear granularity are non-overlapping and their index order is same as time order. \autoref{fig:linear-time} illustrates the linear time granularites hours, days, weeks, months and years, each of which are non-repeating in nature and arranged from past to future. If we consider the chronon [see @aigner2011visualization] as hourly, the time domain with $T$ hours will have $\lfloor T/24\rfloor$ days, $\lfloor T/(24*7)\rfloor$ weeks and so on.

```{r linear-time, echo=FALSE, out.width = "100%", fig.cap="The time domain distributed as linear granularities."}
knitr::include_graphics("Figs/linear-arr.png")
```
In addition to the integer index, each granule could be associated with a textual representation which is called the label. The label can be mapped to index for input and the index can later be mapped to this label for output. When the label set is exactly the integers, the granularities are called "full-integer" labeled granularity, such as in Definition \ref{def:linear}.<!-- Some calendar operations will only work on full-integer labeled granularities, while others can be implemented using more flexible labelings.-->

@@

For example, due to the uneven length of months, we need to use the altering-tick operation to compute months from days.  Let us consider a granularity "pseudomonth" which computes months only for non-leap years by grouping 31 days, and then shrink April, June, September and November by 1 day, and shrink February by 3 days [@Ning2002-tf].

\begin{align*}
\text{pseudomonth} = &\text{Alter}^{12}_{11,-1}(\text{day}, \text{Alter}^{12}_{9,-1}(\text{day}, \\
  &~~\text{Alter}^{12}_{6,-1}(\text{day}, \text{Alter}^{12}_{4,-1)}(\text{day}, \\
  &~~~\text{Alter}^{12}_{(2,-3)}(\text{day}, \text{Group}_{31}(\text{day})))))),
\end{align*}

For more variations of calendar algebra operations, see @Ning2002-tf.

@@

We propose a formalism of cyclic time granularities through the tsibble [@wang2019tsibble] framework of organizing temporal data. A tsibble consists of an index, key and measured variables. An index is a variable with inherent ordering from past to present and a key is a set of variables that define observational units over time. In a tsibble, each observation (row) is uniquely identified by index and key.

A time domain, as defined by @Bettini1998-ed, is essentially a mapping of the index set to the time index of a tsibble. A linear granularity is a mapping of the index set to subsets of the time domain. For example, if the time index of a tsibble is days, then a linear granularity might be weeks, months or years. A bottom granularity is represented by the index of the tsibble. Consistent with the assumption of @Bettini2000-vy, all linear granularities can be generated from the bottom granularity by calendar algebraic operations.

Cyclic time granularities with periodicities can be constructed by relating two linear granularities. The mappings between these linear granularities can be regular (minutes and hours) or irregular (days and months). <!--For example, a regular mapping exists between  (where 60 minutes always add up to 1 hour), while an irregular mapping exists between  (as the number of days per month ranges from 28 to 31).--> A grouping operation in Calendar algebra could lead to a regular mapping, whereas a altering-tick operation could result in an irregular mapping. Cyclic time granularities are referred to as **circular** if mappings are regular and **quasi-circular** if the mapping is irregular. Examples of common circular granularities include hour of the day, and day of the week, whereas examples for quasi-circular granularities can be day of the month or week of the month.

@@

**Example**: Example showing quasi-circular granularities relating two linear granularities each with bottom granularities are visually depicted in a series of slots in \autoref{fig:quasi-circular-example}. Each granule is represented by a box. Two linear granularities $G' = Alter^{2}_{(1,-1)}(BG, Group_3(B))$ and $H' = Alter^{2}_{(1,-2)}(BG, Group_7(B))$ are considered. This implies that $G'$ is made up by shrinking every 1st granule of $Group_3(B)$ by $1$ granule and $H'$ is made up of shrinking every 1st granule of $Group_3(B)$ by $2$ granules. Number of granules of $G'$ and $H'$ in each period of $B$ is 2 but the number of granules of $B$ in each of those granules are different. $Q_{B, G'}$ and $Q_{B, H'}$ are repetitive categorization of time, similar to circular granularities, except that the number of granules of $B$ is not necessarily the same across different granules of $G'$ or $H'$. For $G'$, $T_0 = \{0,1\}$ and $T_1 = \{2, 3, 4\}$. For $H'$, $T_0 = \{0, 1, 2, 3, 4, 5, 6\}$ and $T_1 = \{7, 8, 9, 10, 11\}$, then we will have \autoref{eq:eq4} and  \autoref{eq:circular-gran}.

@@

It can be noted here that if a linear granularity $G'$ is quasi-periodic (Bettini2000-vy) <!--(Definition \autoref{def:quasi})--> with respect to $B$, then \autoref{eq:quasi} can be modified as follows to account for exceptions $E = [e_{begin}, e_{end}]$.<!--(by Definition \autoref{def:quasi}).-->

\begin{equation}\label{eq6}
A_{B, G'}(z) = L(z - \sum_{w=0}^{k-1}\vert T_{w \mod R'} \vert - \sum_{u=0}^{e_k}\vert E_{u} \vert)
\end{equation}

for $z \in T_k$ and $e_k$ is the number of exceptions in $\bigcup\limits_{w = 0}^{k} T_w$ and $E_{u} = [e_{begin}(u), e_{end}(u)]$.

@@

Periodic and Quasi-periodic behavior can be defined by a repeating pattern, except for a finite number of granules that can be seen as discontinuity points in the granularity in case of quasi-periodic behavior. Aperiodic time granularities are the ones which can not be specified as a periodic repetition of a pattern of granules. Most public holidays repeat every year, but there is no finite (or reasonably small) period within which their behavior remains constant. A classic example can be that of Easter, whose dates repeat only after 5,700,000 years. U.S labor day is the first Monday in September and U.S. Memorial Day is the last day in May. In Australia, if a standard public holiday falls on a weekend, a substitute public holiday will sometimes be observed on the first non-weekend day (usually Monday) after the weekend. Examples of aperiodic granularity may also include school holidays or a scheduling event that might cover the first and third Monday of the month between June and October, except for state holidays. All of these are recurring events, but with non-periodic patterns. As such, plausible $P_i$ from \autoref{tab:quasi} could be possibly infinite for aperiodic granularities.

Aperiodic cyclic granularities are defined using aperiodic linear granularities. Consider n aperiodic linear granularities $M_i$  $\forall \{i\in{1, 2, \dots, n\}}$, that is, none of them can be expressed as periodic or quasi-periodic with respect to the bottom granularity. Further assume that $B \trianglelefteq M_i$ $\forall \{i\in{1, 2, \dots, n\}}$. Then according to Definition \ref{def:groupsinto}, for each index $j$ there exists a (possibly infinite) subset $T_{\{i_j\}}$ of the integers such that $M_{i}(j) = \bigcup_{z \in T_{i_j}}B(z)$ and suppose $M = \bigcup_{i=1}^{n}M_{i}$ is formed by collecting the granules of $\{M_1, M_2, \dots, M_n\}$. Here, index ${\{i_j\}}$ stands for the $j^{\text{th}}$ granule of the $i^{\text{th}}$ linear aperiodic granularity.

@@

**Example**: \autoref{fig:aperiodic-example} shows two aperiodic linear granularity $M_1$ and $M_2$. Each granule is represented by a box. The diagram also illustrates that the granules that overlap share elements from the underlying time domain. The first slot in the diagram shows the index set and $A_{B, M}$ consists of three categories, $L(1)$ corresponding to the first aperiodic event $M_1$, $L(2)$ corresponding to aperiodic event $M_2$ and the $L(0)$ corresponding to no event/aperiodic linear granularity.

**Discussion**: Here, $M_1$ can be an aperiodic linear granularity denoting block leaves during Christmas, $M_2$ can be the one denoting Easter, $M_n$ can be Lab our day and M can be conceived as the aperiodic linear granularity denoting any holidays. The number of categories obtained coincides with the number of aperiodic linear granularities considered and an additional category representing no event.

@@

In \autoref{sec:cyclic-gran}, we discussed how we can define cyclic granularities by extending the framework of linear granularities defined in @Bettini1998-ed. In this section, we will see how to obtain cyclic granularities through an algebraic representation of other cyclic granularities. This is similar to calendar algebra in @Ning2002-tf, where linear granularities are generated from the bottom linear granularity or from those that are already generated. Since, our method caters to computation of cyclic granularities, we shall refer to it as "cyclic calendar algebra".

@@

<!-- some discussions on primary units of calendar -->
**Example:** So far we have used example of cyclic granularities from Gregorian calendar as it is the most widely used calendar. But it is far from being the only one. <!--All calendars fall under three types - solar, lunar or lunisolar/solilunar-->. The day is the basic unit of time underlying most calendars [@Reingold2001-kf]. Various calendars, however, use different conventions to structure days into larger units: weeks, months, years and cycle of years. The French revolutionary calendar divided each day into 10 "hours", each "hour" into 100 "minutes" and each "minute" into 100 "seconds". Nevertheless, for any calendar a hierarchy table can be defined. For example, in Mayan calendar, one day was referred to as 1 kin and the calendar was structured such that 1 kin = 1 day; 1 uinal = 20 kin; 1 tun  = 18 uinal; 1 katun = 20 tun and 1 baktun = 20 katun

@@

Let us revisit Gregorian calendar for addressing this case. Suppose we have a hierarchy table using some linear granularties from Gregorian calendar. Since months consists of unequal number of days, any linear granularity with higher order than months will also have unequal number of days. This is an example of a hierarchy structure which has both circular and quasi-circular single-order-up granularities. The single-order-up granularity day-of-month is quasi-circular. Any single-order-up granularities which are formed by linear granularities below days are circular. Similarly, all single-order-up granularities which are formed using linear granularities with orders higher than months are also circular.

@@

Generally speaking, to keep it simple, we consider the case of only one quasi-circular single order-up granularity in the hierarchy table. Any multiple order-up quasi-circular granularity $C_{l, m}(z)$ can then be considered to be a function of two single order-up granularities $C_{l,m'}(z)$ and $C_{m',m}(z)$. Two different approaches need to be employed for the following cases:

>>>>

The first condition in Definition \ref{def:linear} implies that the granules in a linear granularity are non-overlapping and their index order is same as time order. \autoref{fig:linear-time} illustrates the linear time granularites hours, days, weeks, months and years, each of which are non-repeating in nature and arranged from past to future. If we consider $T$ hours, we will have $\lfloor T/24\rfloor$ days, $\lfloor T/(24*7)\rfloor$ weeks and so on.

```{r linear-time, echo=FALSE, out.width = "100%", fig.cap="The time domain distributed as linear granularities."}
knitr::include_graphics("Figs/linear-arr.png")
```
In addition to the integer index, each granule could be associated with a textual representation which is called the label.<!--The label can be mapped to index for input and the index can later be mapped to this label for output.--> When the label set is exactly the integers, the granularities are called "full-integer" labeled granularity, such as in Definition \ref{def:linear}.<!-- Some calendar operations will only work on full-integer labeled granularities, while others can be implemented using more flexible labelings.-->

@@

For example, due to the uneven length of months, we need to use the altering-tick operation to compute months from days. We can consider a granularity "pseudomonth" which computes months only for non-leap years by grouping 31 days, and then shrink April (4), June (6), September (9) and November (11) by 1 day, and shrink February (2) by 3 days [@Ning2002-tf]. For more variations of calendar algebra operations, see @Ning2002-tf.

<!-- CONTENT CUT -->
<!-- \begin{align*} -->
<!-- \text{pseudomonth} = &\text{Alter}^{12}_{11,-1}(\text{day}, \text{Alter}^{12}_{9,-1}(\text{day}, \\ -->
<!--   &~~\text{Alter}^{12}_{6,-1}(\text{day}, \text{Alter}^{12}_{4,-1)}(\text{day}, \\ -->
<!--   &~~~\text{Alter}^{12}_{(2,-3)}(\text{day}, \text{Group}_{31}(\text{day})))))), -->
<!-- \end{align*} -->

@@

We propose a formalism of cyclic time granularities through the tsibble [@wang2019tsibble] framework of organizing temporal data. A tsibble consists of an index, key and measured variables. An index is a variable with inherent ordering from past to present and a key is a set of variables that define observational units over time. <!--In a tsibble, each observation (row) is uniquely identified by index and key.--> A time domain, as defined by @Bettini1998-ed, is essentially a mapping of the index set to the time index of a tsibble. A linear granularity is a mapping of the index set to subsets of the time domain. For example, if the time index of a tsibble is days, then a linear granularity might be weeks, months or years. A bottom granularity is represented by the index of the tsibble.<!--Consistent with the assumption of @Bettini2000-vy, all linear granularities can be generated from the bottom granularity by calendar algebraic operations.-->

Cyclic granularities can be constructed by relating two linear granularities. The mappings between these linear granularities could be _regular_ with every granule of coarser linear granularity being made up of same number of granules of finer linear granularity or _irregular_. <!--For example, a regular mapping exists between  (where 60 minutes always add up to 1 hour), while an irregular mapping exists between  (as the number of days per month ranges from 28 to 31). A grouping operation in Calendar algebra could lead to a regular mapping, whereas a altering-tick operation could result in an irregular mapping.--> Cyclic time granularities are referred to as **circular** if mappings are regular and **quasi-circular** if the mappings are irregular. Examples of common circular granularities include hour of the day, and day of the week, whereas examples for quasi-circular granularities can be day of the month or week of the month.

@@

**Example**: Example showing quasi-circular granularities relating two linear granularities each with bottom granularities are visually depicted in a series of slots in \autoref{fig:quasi-circular-example}. Each granule is represented by a box. Two linear granularities $G' = Alter^{2}_{(1,-1)}(BG, Group_3(B))$ and $H' = Alter^{2}_{(1,-2)}(BG, Group_7(B))$ are considered. <!--This implies that $G'$ is made up by shrinking every 1st granule of $Group_3(B)$ by $1$ granule and $H'$ is made up of shrinking every 1st granule of $Group_3(B)$ by $2$ granules. Number of granules of $G'$ and $H'$ in each period of $B$ is 2 but the number of granules of $B$ in each of those granules are different.--> $Q_{B, G'}$ and $Q_{B, H'}$ are repetitive categorization of time, similar to circular granularities, except that the number of granules of $B$ is not necessarily the same across different granules of $G'$ or $H'$. For $G'$, $T_0 = \{0,1\}$ and $T_1 = \{2, 3, 4\}$. For $H'$, $T_0 = \{0, 1, 2, 3, 4, 5, 6\}$ and $T_1 = \{7, 8, 9, 10, 11\}$, then we will have \autoref{eq:eq4} and  \autoref{eq:circular-gran}.

@@

<!-- CONTENT CUT -->
<!-- It can be noted here that if a linear granularity $G'$ is quasi-periodic (Bettini2000-vy) <!--(Definition \autoref{def:quasi}) with respect to $B$, then \autoref{eq:quasi} can be modified as follows to account for exceptions $E = [e_{begin}, e_{end}]$.<!--(by Definition \autoref{def:quasi}).-->

<!-- \begin{equation}\label{eq6} -->
<!-- A_{B, G'}(z) = L(z - \sum_{w=0}^{k-1}\vert T_{w \mod R'} \vert - \sum_{u=0}^{e_k}\vert E_{u} \vert) -->
<!-- \end{equation} -->

<!-- for $z \in T_k$ and $e_k$ is the number of exceptions in $\bigcup\limits_{w = 0}^{k} T_w$ and $E_{u} = [e_{begin}(u), e_{end}(u)]$. -->

@@

<!--Periodic and Quasi-periodic behavior can be defined by a repeating pattern except for a finite number of granules that can be seen as discontinuity points in the granularity in case of quasi-periodic behavior.--> Aperiodic time granularities are the ones which can not be specified as a periodic repetition of a pattern of granules. Most public holidays repeat every year, but there is no finite (or reasonably small) period within which their behavior remains constant. A classic example can be that of Easter, whose dates repeat only after 5,700,000 years.<!--U.S labor day is the first Monday in September and U.S. Memorial Day is the last day in May. --> In Australia, if a standard public holiday falls on a weekend, a substitute public holiday will sometimes be observed on the first non-weekend day (usually Monday) after the weekend. Examples of aperiodic granularity may also include school holidays or a scheduling event.<!--that might cover the first and third Monday of the month between June and October, except for state holidays.--> All of these are recurring events, but with non-periodic patterns. As such, plausible $P_i$ from \autoref{tab:quasi} could be possibly infinite for aperiodic granularities.

Aperiodic cyclic granularities are defined using aperiodic linear granularities. Consider n aperiodic linear granularities $M_i$  $\forall \{i\in{1, 2, \dots, n\}}$. <!--that is, none of them can be expressed as periodic or quasi-periodic with respect to the bottom granularity.--> Further assume that $B \trianglelefteq M_i$ $\forall \{i\in{1, 2, \dots, n\}}$. Then according to Definition \ref{def:groupsinto}, for each index $j$ there exists a (possibly infinite) subset $T_{\{i_j\}}$ of the integers such that $M_{i}(j) = \bigcup_{z \in T_{i_j}}B(z)$ and suppose $M = \bigcup_{i=1}^{n}M_{i}$ is formed by collecting the granules of $\{M_1, M_2, \dots, M_n\}$. Here, index ${\{i_j\}}$ stands for the $j^{\text{th}}$ granule of the $i^{\text{th}}$ linear aperiodic granularity.

@@

**Example**: \autoref{fig:aperiodic-example} shows two aperiodic linear granularity $M_1$ and $M_2$. <!--Each granule is represented by a box. The diagram also illustrates that the granules that overlap share elements from the underlying time domain. -->The first slot in the diagram shows the index set and $A_{B, M}$ consists of three categories, $L(1)$ corresponding to the first aperiodic event $M_1$, $L(2)$ corresponding to aperiodic event $M_2$ and the $L(0)$ corresponding to no event/aperiodic linear granularity.

**Discussion**: Here, $M_1$ can be an aperiodic linear granularity denoting block leaves during Christmas, $M_2$ can be the one denoting Easter, $M_n$ can be Labor day and M can be conceived as the aperiodic linear granularity denoting any holidays. <!--The number of categories obtained coincides with the number of aperiodic linear granularities considered and an additional category representing no event.-->

@@

In \autoref{sec:cyclic-gran}, we discussed how we can define cyclic granularities by extending the framework of linear granularities defined in @Bettini1998-ed. In this section, we will see how to obtain cyclic granularities through an algebraic representation of other cyclic granularities <!--This is similar to calendar algebra in @Ning2002-tf, where linear granularities are generated from the bottom linear granularity or from those that are already generated. Since, our method caters to computation of cyclic granularities, we shall refer to it as--> through cyclic calendar algebra operations.

@@

<!-- some discussions on primary units of calendar -->
**Example:** So far we have used example of cyclic granularities from Gregorian calendar as it is the most widely used calendar. But it is far from being the only one. <!--All calendars fall under three types - solar, lunar or lunisolar/solilunar-->The day is the basic unit of time underlying most calendars [@Reingold2001-kf]. Various calendars, however, use different conventions to structure days into larger units: weeks, months, years and cycle of years. The French revolutionary calendar divided each day into 10 "hours", each "hour" into 100 "minutes" and each "minute" into 100 "seconds". Nevertheless, for any calendar a hierarchy table can be defined. For example, in Mayan calendar, one day was referred to as 1 kin and the calendar was structured such that 1 kin = 1 day; 1 uinal = 20 kin; 1 tun  = 18 uinal; 1 katun = 20 tun and 1 baktun = 20 katun

@@

Let us revisit Gregorian calendar for addressing this case. Suppose we have a hierarchy table using some linear granularties from Gregorian calendar. Since months consists of unequal number of days, any linear granularity with higher order than months will also have unequal number of days. This is an example of a hierarchy structure which has both circular and quasi-circular single-order-up granularities. The single-order-up granularity day-of-month is quasi-circular. <!--Any single-order-up granularities which are formed by linear granularities below days are circular. Similarly, all single-order-up granularities which are formed using linear granularities with orders higher than months are also circular.-->

@@

To keep it simple, we consider the case of only one quasi-circular single order-up granularity in the hierarchy table. Any multiple order-up quasi-circular granularity $C_{l, m}(z)$ can then be considered to be a function of two single order-up granularities $C_{l,m'}(z)$ and $C_{m',m}(z)$. Two different approaches need to be employed for the following cases:
